---

output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab report #3 - instructions

Follow the instructions posted at https://ds202-at-isu.github.io/labs.html for the lab assignment. The work is meant to be finished during the lab time, but you have time until Monday evening to polish things.

Include your answers in this document (Rmd file). Make sure that it knits properly (into the md file). Upload both the Rmd and the md file to your repository.  

All submissions to the github repo will be automatically uploaded for grading once the due date is passed.
Submit a link to your repository on Canvas (only one submission per team) to signal to the instructors that you are done with your submission.

# Lab 3: Avenger's Peril

## As a team

Extract from the data below two data sets in long form `deaths` and `returns`

```{r}
av <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/avengers/avengers.csv", stringsAsFactors = FALSE)
head(av)
```

Get the data into a format where the five columns for Death[1-5] are replaced by two columns: Time, and Death. Time should be a number between 1 and 5 (look into the function `parse_number`); Death is a categorical variables with values "yes", "no" and "". Call the resulting data set `deaths`. 
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)

# Transforming Death columns into long format
deaths <- av %>%
  pivot_longer(cols = starts_with("Death"), names_to = "Time", values_to = "Death") %>%
  mutate(Time = parse_number(Time), Death = replace_na(Death, ""))
head(deaths)

```


Similarly, deal with the returns of characters.
```{r}
# Transforming Return columns into long format
returns <- av %>%
  pivot_longer(cols = starts_with("Return"), names_to = "Time", values_to = "Return") %>%
  mutate(Time = parse_number(Time), Return = replace_na(Return, ""))
head(returns)

```


Based on these datasets calculate the average number of deaths an Avenger suffers.
```{r}
average_deaths <- deaths %>%
  filter(Death == "YES") %>%
  group_by(`Name.Alias`) %>%
  summarize(num_deaths = n()) %>%
  summarize(avg_deaths = mean(num_deaths))
average_deaths

```


## Individually

For each team member, copy this part of the report. 

Each team member picks one of the statements in the FiveThirtyEight [analysis](https://fivethirtyeight.com/features/avengers-death-comics-age-of-ultron/) and fact checks it based on the data. Use dplyr functionality whenever possible.

### Keenan Jacobs:
### FiveThirtyEight Statement

> "I counted 89 total deaths — some unlucky Avengers are basically Meat Loaf with an E-ZPass — and on 57 occasions the individual made a comeback."

### Code to Fact-Check

```{r}
# Counting total deaths
total_deaths <- deaths %>%
  filter(Death == "YES") %>%
  summarize(total_deaths = n())

# Counting total returns
total_returns <- returns %>%
  filter(Return == "YES") %>%
  summarize(total_returns = n())

# Display both totals
total_deaths
total_returns
```

### My Answer

The data confirms the statement from FiveThirtyEight regarding the Avengers' mortality and comebacks. We found exactly 89 deaths and 57 returns among the characters, aligning with the claim that "89 total deaths" and "57 occasions the individual made a comeback."

---
### Muhammad Raham Saleem:

### FiveThirtyEight Statement


> But you can only tempt death so many times. There’s a 2-in-3 chance that a member of the Avengers returned from their first stint in the afterlife, but only a 50 percent chance they recovered from a second or third death.

### Code
```{r}
# Ensure consistent lowercase values in Death and Return columns
deaths <- deaths %>%
  mutate(Death = tolower(Death))

returns <- returns %>%
  mutate(Return = tolower(Return))

# Merge the 'deaths' and 'returns' datasets based on Name.Alias and Time
death_return <- left_join(deaths, returns, by = c("Name.Alias", "Time"), relationship = "many-to-many")

# Calculate the return probability for the first three death instances only
return_probabilities <- death_return %>%
  filter(Death == "yes", Time <= 3) %>%  # Only consider rows where Death is "yes" and Time is 1, 2, or 3
  group_by(Time) %>%
  summarize(
    total_deaths = n(),                    # Count of total deaths at each time (e.g., first death, second death)
    total_returns = sum(Return == "yes"),   # Count of returns after each death
    return_probability = total_returns / total_deaths # Probability of return
  )

# Display the calculated return probabilities for the first three deaths
print("Return probabilities for the first three death instances (Time):")
return_probabilities



```

Make sure to include the code to derive the (numeric) fact for the statement

### Include your answer

First Death:

The return probability after the first death is approximately 45.5% (0.455), which is lower than the expected 2-in-3 chance (66.7%).
This means that, contrary to the claim, Avengers have less than a 50% chance of returning after their first death based on this dataset.
Second Death:

The return probability after the second death is 32.0% (0.32), which is below the expected 50%.
This suggests that fewer Avengers return after their second death than claimed.
Third Death:

The return probability after the third death is 50% (0.5), which aligns with the claim of a 50% chance of returning after a third death.
Fourth and Fifth Deaths:


After the first death, the return probability is 45.5%, not the claimed 66.7%.
After the second death, the return probability is 32%, lower than the expected 50%.
After the third death, the return probability does align with 50%, which is consistent with the claim.
These findings suggest that the likelihood of returning from death decreases over time and is generally lower than the quoted statistics in the claim.


###Bradyn Weaver

### FiveThirtyEight Statement

> Given the Avengers’ 53 years in operation and overall mortality rate, fans of the comics can expect one current or former member to die every seven months or so, with a permanent death occurring once every 20 months.


```{r}
#For calculating the average months per death:

total_deaths <- sum(av$Death1 == "YES", na.rm = TRUE)
permanent_deaths <- sum(av$Return1 == "NO", na.rm = TRUE)

# Assuming the dataset covers 53 years of Avengers history:
years_in_operation <- 53

# Calculate the average time between any death and permanent deaths
average_months_per_death <- (years_in_operation * 12) / total_deaths
average_months_per_permanent_death <- (years_in_operation * 12) / permanent_deaths

# Print results
cat("Average time between any death:", average_months_per_death, "months\n")
cat("Average time between permanent deaths:", average_months_per_permanent_death, "months\n")

```

### Include your answer

Based on the dataset analysis above, it appears that FiveThirtyEight may have been off by 2 months for their average time between any death estimate. It also appears they were off by 7 months for their permanent death rate estimate.


### Meghasyam Peddireddy

### FiveThirtyEight Statement

> "If a character is killed but then secretly hidden away in a stasis tube, they died."

### Code to Fact-Check

```{r}
# Ensuring consistent values in Death and Return columns
deaths <- deaths %>%
  mutate(Death = tolower(Death))

returns <- returns %>%
  mutate(Return = tolower(Return))

# Merging deaths and returns datasets to check cases where a character 'died' but later 'returned'
death_stasis_check <- deaths %>%
  inner_join(returns, by = c("Name.Alias", "Time")) %>%
  filter(Death == "yes" & Return == "yes")

# Counting the number of such "stasis tube" instances
stasis_count <- nrow(death_stasis_check)

# Displaying the result
death_stasis_check
stasis_count

```


### My Answer
Based on the result from the `death_stasis_check` table, we identified **`r stasis_count`** instances where characters were marked as "dead" but later "returned." This supports the FiveThirtyEight statement that Avengers who are believed to be dead may not always remain so, with some effectively being "hidden away in a stasis tube" or similar scenarios that enable their return. This finding aligns with the article's claim, illustrating that in the Marvel universe, death is often temporary, and characters frequently find ways to reappear after seeming to perish.


